---
title: "Assignemnt 2: Term structure and expectations"
output: pdf_document
Author: Riccardo_DC
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pandoc)
```



```{r p0, message=FALSE, echo=FALSE}
# Clear the variables
rm(list = ls())

# Install packages
packages <- c("matlib", "rmarkdown", "tinytex","plotly","rgl","knitr","pandoc")
new.packages <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(new.packages)) install.packages(new.packages)
invisible(lapply(packages, library, character.only = TRUE))
# Load packages
library(matlib)
library(scatterplot3d)
library(plotly)
library(rgl)
library(knitr)
options(digits = 15) 
library(pandoc)
```


## Exercise 1:

```{r p1, message=FALSE, echo=FALSE}
# Clear the variables
rm(list = ls())
library(matlib, help, pos = 2, lib.loc = NULL)
phi       <- 0.3

prob      <- matrix(c(phi, 1 - phi), 2, 2)
prob[, 2] <- c(1 - phi, phi)
gamma  <- 2
beta    <- .96
h       <- 1.05
l       <- .97
state   <- c(h, l) ^ - gamma

```

The exogenous endowment shock
is a two states Markov process, defined as:

$$\frac{\Omega_{t}}{\Omega_{t-1}}=\omega_{t}\quad 
  with \; \omega_{t}\epsilon (h,l)$$

With the following symmetric transition matrix:

$$\Pi = \left[\begin{matrix}\phi & (1-\phi) \\ 
                            (1-\phi) & \phi \\ \end{matrix}\right]$$

Let $\beta = 0.96$, $\gamma=2$, $h=1.05$, $l=0.97$ and $\phi=0.3$.

Knowing this, we cannot use the same 4-cases framework 
The probability of finding ourselves in state h or state l now depends on the state of the economy in the
previous period.

Thus, we will have two different yield curves, conditional on the state
of the Markov process we are starting with at time t. 

We will generate a function that takes as inputs the two status of the markow chain $(high, low)$, the transition matrix $\Pi$ and the number of
step $(n)$ to itereate the Markov process so the algorithm is scalable. The algorithm follow these steps:

1. iteration count <- 1
2. $$States = \Pi * \left[\begin{matrix} high \\ low \\ \end{matrix}\right] $$
3. $$yieds = - \ln(\beta) - \frac{1}{count}\ln(States)$$
4. adding yields to output
5. $$ \Pi = \Pi * \Pi $$
6. $$ state = state * state^T$$
7. restart at point 1 until $count = n$
```{r function, echo= TRUE, include = FALSE}

yield_mcmc <- function(n, prob, state) {
    out <- list()
    for (count in 1:n) {
        state <-   prob %*% state
        y <-  -(1 / count) * log(state) - log(beta)
        out <- cbind(out, y)
        prob  <-  prob   %*% prob
        state <- state  %*% t(state)
    }
    return(out)
}

```



## Iteration for 1 period
We wanted also to have a measure of the short-term interest rate in 
period t. In the real world, this is the overnight interest rate, set
by the Central Bank. 
Let's look now at the output:

```{r p2, message=FALSE, echo=FALSE}

yieldcurve <- yield_mcmc(1, prob, state) 
yieldcurve <- as.matrix(yieldcurve)
out <- matrix(NA, nrow = 2, ncol = 1)

out[, 1] <- c(as.numeric(yieldcurve[1, 1]), as.numeric(yieldcurve[2,1]))
colnames(out) <- c("t=1")
rownames(out) <- c("h", "l")
kable(out)

plot(NULL, xlim = c(0, length(yieldcurve[1,])+1), ylim = c(-1,10), type = "n", xlab='Maturity', ylab="Rate")
points( xy.coords(1, as.numeric(yieldcurve[1,])*100), type = 'p', col = '#0077ff')
points( xy.coords(1, as.numeric(yieldcurve[2,])*100), type = 'p', col = '#ff0080')
legend("topright",legend=c("High Status", "Low Status"),
       col=c("#0077ff", "#ff0080"), lty= 1:1, cex=1)
title("Yield curve with 1 period")
abline(h = 0, col = "black", lty="dashed")


```

At time 1 the economy can be at state high or a state low, if it is in high status the yield curve starts 
at $2.49\%$ otherwise the economy is in low state then the yield curve starts at $8.88\%$.

We observe that even if the curve predicts a recovery in the long term, does
not return to the first-period yield level. We believe this is because 
in the starting period the state $h$ is certain, while in the second it is 
only slightly more likely than a recession, thus the yield is lower.



## Introducing two time (n = 2)


```{r p3, message=FALSE, echo=FALSE}
yieldcurve <- yield_mcmc(2, prob, state)
yieldcurve <- as.matrix(yieldcurve)
yieldcurve_2 <- matrix(yieldcurve[,c(2,3)], nrow = 2, ncol = 2)
yieldcurve_3 <- c(phi * as.numeric(yieldcurve_2[1, 1]) 
+ (1 - phi) * as.numeric(yieldcurve_2[1, 2]),
(1 - phi) * as.numeric(yieldcurve_2[2, 2]) + phi 
* as.numeric(yieldcurve_2[2, 1]))
out <- matrix(NA, nrow = 2, ncol = 2)
colnames(out) <- c("t=1","t=2")
rownames(out) <- c("h", "l")
out[, 1] <- c(as.numeric(yieldcurve[1, 1]), as.numeric(yieldcurve[2, 1]))
out[, 2] <- yieldcurve_3
kable(out)

plot(NULL, xlim = c(0, 2), 
ylim = c(-1,10), type = "n", xlab = 'Maturity', ylab = "Rate")
a <- c(0, 0)
a[1] <- as.numeric(yieldcurve[1, 1]) * 100
a[2] <- as.numeric(yieldcurve_3[1]) * 100
lines(a, type = 'b', col = '#0077ff')

b <- c(0, 0)
b[1] <- as.numeric(yieldcurve[2, 1]) * 100
b[2] <- as.numeric(yieldcurve_3[2]) * 100
lines(b, type = 'b', col = '#ff0080')



legend("topright",legend=c("High Status", "Low Status"),
       col=c("#0077ff", "#ff0080"), lty= 1:1, cex=1)
title("Yield curve with 1 period")
abline(h = 0, col = "black", lty="dashed")
```


In this case, the yield curve starts as in the previous case at time t=1, in which the economy
can be in a good state or in a bad state as mentioned above. Conditioning to state of the economy
at t= 1 and itereting the markov process to generate new transition probability ($n=2$), we get the yields
curve at time t=2.

According to the red line rappresting the low state of the economy, the slope is negative indicating a recession.
While the high  status yield it has a positive slope coeficient. Both yield converges around 6\% this is probably due 
to the convergence of the transition probability as the number of iteration becames greater.


# Exercise 2

```{r setup2, include=FALSE, echo= FALSE}
library(ecb)
library(xts)
library(ggplot2)
filter1 = list(startPeriod = "2020-07", endPeriod = "2022-07") #filtering time

#import data for spot rates
sp          <-  matrix(0, nrow = 535, ncol = 10)
keys        <-  "YC.B.U2.EUR.4F.G_N_A.SV_C_YM.SR_"
colnamess   <- c()
for (count in 1:10) {
    keys_count          <- paste(keys, count, sep = "")
    keys_count          <- paste(keys_count, "Y", sep = "")
    data                <- get_data(key = keys_count,filter = filter1)
    data$obstime        <- convert_dates(data$obstime)
    colname             <- paste(count, "Y")
    sp[ ,count]          <- data$obsvalue
    colnamess           <- cbind(colnamess, colname)
}
colnames(sp)            <- colnamess
#sp      <- xts(sp,order.by = data$obstime)

#import data for forward price 
fw          <-  matrix(NA, nrow = 534, ncol = 10)
keys        <-  "YC.B.U2.EUR.4F.G_N_A.SV_C_YM.IF_"
colnamess   <- c()
for (count in 1:10) {
    keys_count          <- paste(keys, count, sep = "")
    keys_count          <- paste(keys_count, "Y", sep = "")
    data                <- get_data(key = keys_count,filter = filter1)
    data$obstime        <- convert_dates(data$obstime)
    colname             <- paste(count, "Y")
    fw[,count]          <- data$obsvalue
    colnamess           <- cbind(colnamess, colname)
}
colnames(fw)            <- colnamess
#fw      <- xts(fw,order.by = data$obstime)

year <- 1:10
data <- matrix(NA, nrow=10, ncol=2)
data[ ,1]    <- sp[1,]
data[ ,2]    <- fw[1,]

colnames(data) <- c("Spot", "Forward")
rownames(data) <- colnamess
data <- data.frame(data)
```
# Plot yield curves 2020
```{r plot10, echo=FALSE, include=TRUE}
#plot first
plot(NULL, xlim = c(0, nrow(data)), ylim = c(-1,2), type = "n", xaxt = 'n', xlab='Maturity', ylab="Rate")
lines( data$Spot, type = 'b', col = '#0077ff')
lines(data$Forward, type = 'b', col = '#ff00a2')
abline(h = 0, col = "black", lty="dashed")
axis(1, at = 1:nrow(data), labels = rownames(data))
legend("topright",legend=c("Spot", "Forward"),
       col=c("#0077ff", "#ff00a2"), lty= 1:1, cex=1)
title("Yield curve at 2020")
```
The plot above represents the spot and forward yield curve in July 2020. The spot yield is entirely 
in the negative area of interest rate due to the APP of the ECB. Even the forward curve is negative
until the 8Y maturity. Indeed, In addition, the ECB enlarged the APP program to face the pandemic in the EU countries, 
lowering the yield curve in the short maturity.

The curves are upward-sloping after 3Y, and the forward rate premium becomes higher as
maturity increases. Suggesting that agents had been expecting the recovery of consumption at the pre-covid level,
a tightening in the monetary policy, and probably higher inflation rates.

Regarding the maturity before 3Y, the forward rate is above the spot rate, and both have negative slopes. 
Leading to the belief that the agents had incorporated a recession in the 3 years ahead. As Engstrom and Sharpe (2018)
find that a near-term negative spread may only predict recessions because it reflects 
the market's expectation that a contracting economy will induce the central bank to lower its policy rate.
Indeed, the nominal interest rate is driven by the following:

* $E(Inflation rates)$ -> inflation risk premium
* $E(Monetary policy)$ -> real rate risk premium

Hence, if investors see higher odds of a recession, the long-term inflation risk premium in Treasury bonds will fall.
In contrast, an increase in the recession probability would increase the real rate risk premium asked by agents. 
One interpretation can be that if investors see a greater risk of recession, they will attribute a higher value 
to short-term assets that they can quickly liquidate to finance spending on goods and services.
It seems reasonable because the covid had already broken out in July, and the prices incorporated 
the expectation of future covid measures that hinder consumption.

# Plot yield curves 2022

```{r plot20, echo=FALSE, include=TRUE}
data[ ,1]    <- sp[515,]
data[ ,2]    <- fw[514,]

colnames(data) <- c("Spot", "Forward")
rownames(data) <- colnamess
data <- data.frame(data)

#plot first
plot(NULL, xlim = c(0, nrow(data)), ylim = c(-0.5,2.5), type = "n", xaxt = 'n', xlab='Maturity', ylab="Rate")
lines( data$Spot, type = 'b', col = '#0077ff')
lines(data$Forward, type = 'b', col = '#ff00a2')
abline(h = 0, col = "black", lty="dashed")
axis(1, at = 1:nrow(data), labels = rownames(data))
legend("bottomright",legend=c("Spot", "Forward"),
       col=c("#0077ff", "#ff00a2"), lty= 1:1, cex=1)
title("Yield curve at 2022")
```

The plot above represents the forward and spot yield curves for different maturity quoted in July 2022.
Compared to the previous plot, the forward yield curve is steeper. Thus the forward risk premium is higher for each 
maturity than the last plot. A possible interpretation is that the agents had expected higher inflation, an increase in monetary tightening, asking for a higher risk premium.
Regarding the slopes of the two curves, the forward rate slope is positive, greater in absolute values compared to one year before
ar each maturity. While the slope of the curves are both positive positive.
The slopes and the forward risk premium are positive, this hints that agents believe in a positive state of the economy 
in the long run.